Retrofitting Effect Handlers onto OCaml

### 2020-11-18 retro-concurrency.tex
Updates

%\vspace{-2mm}
\subsection{Exceptions}

@@

%\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

@@

Next, we have the variable-sized area for the OCaml frames, known as the red zone.
In order to keep fibers small, this area is initially 16 words in length.
When the stack pointer |rsp| becomes less than the stack threshold
(maintained in the |Caml_state| table), the stack is said to have overflown.
On stack overflow, we copy the
whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.
In our evaluation of real world OCaml programs ((\S\ref{sec:eval}) we observed that most function
calls are to leaf functions with a small frame size. We introduce a small red zone at the top of the stack to take advantage of this fact. The
compiler elides the stack overflow check for leaf functions whose frame size is
less than the size of the red zone.

@@

%\vspace{-2mm}
\subsection{External calls and callbacks}

@@

%\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\subsection{Stack unwinding}

@@

\subsection{Garbage collection}

@@

\section{Evaluation}
\label{sec:eval}

@@

%\vspace{-3mm}
\subsection{No effects benchmarks}

@@

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower} than |stock|. The outliers (on either ends) were due to the difference
in the allocator and the GC between stock and Multicore OCaml. The biggest
impact was the increase in the code size for OCaml code due to stack overflow
checks. Compared to |stock|, the code size for OCaml code is 19\% more for |MC|
and |MC+RedZone32|, and 30\% more for |MC+RedZone0|. The result shows that our
red zone is effective at reducing the code size, but further work is required
to bring it closer to |stock|.

@@

%\vspace{-3mm}
\subsection{No perform benchmarks}

@@

%\vspace{-3mm}
\subsection{Concurrent benchmarks}

@@

%\vspace{-3.5mm}
\section{Related Work}
\label{sec:related}

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is not quite sufficient for resource
clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
from non-returning ones, and resource cleanup is only necessary during
non-returning exits.
Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety.

Sivaramakrishnan et al.~\cite{TFP17} describe the interaction of effect
handlers and asynchronous exceptions. This is orthogonal to the contributions
of this paper. Our focus is the compiler and runtime system support for
implementing effect handlers.

\section{Conclusions}
\label{sec:conc}

We have presented the design and implementation of effect handlers
aimed at retrofitting them onto the large body of existing OCaml
code.  Our design walks the tightrope of maintaining
compatibility (e.g. for profiling and debuggers and negligible overheads for existing
programs) and also demonstrating good performance and usability for new OCaml
code that takes advantage of effect handlers. Our implementation in Multicore OCaml
deviates from previous categorisations~\cite{Farvardin20} and our evaluation shows
that we have achieved our goal of balancing compatibility with the introduction
of effect handlers for mainstream OCaml programming.

@@

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the performance
of macro benchmarks that do not use effects. The aim is to quantify the impact
of the addition of effect handlers to the language on code that does not use
effect handlers. The |MC| is Multicore OCaml with the default red zone size (16
words). |MC+RedZone0| uses red zone size of 0; every OCaml function has a stack
overflow check. |MC+RedZone32| uses a red zone size of 32 words.

>>>>

\vspace{-2mm}
\subsection{Exceptions}

@@

\vspace{-2mm}
\subsection{Heap-allocated fibers}
\label{sec:hafibers}

@@

Next, we have the variable-sized area for the OCaml frames. In order to keep
fibers small, this area is initially 16 words in length. When the stack pointer
|rsp| becomes less than the \emph{stack threshold} (maintained in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks into the function prologue of OCaml functions.

In our evaluation of real world OCaml programs (\S\ref{sec:eval}), we observed
that most function calls are to leaf functions with small frame sizes. Can we
eliminate the stack overflow checks for these functions? To this end, we
introduce a small, fixed-sized red zone at the top of the stack. The compiler
elides the stack overflow check for leaf functions whose frame size is less
than the size of the red zone. The default size of the red zone in Multicore
OCaml is 16 words.

@@

\vspace{-2mm}
\subsection{External calls and callbacks}

@@

\vspace{-2mm}
\subsection{Effect handlers}
\label{sec:effimpl}

@@

\vspace{-3mm}
\subsection{Stack unwinding}

@@

\vspace{-2mm}
\subsection{Garbage collection}

@@

\vspace{-2mm}
\section{Evaluation}
\label{sec:eval}

@@

\vspace{-3mm}
\subsection{No effects benchmarks}

@@

The complete results are presented in the supplementary material, which we
summarise here. On average (geometric mean of the normalized values against
|stock| as the baseline), the multicore variants were \textbf{less than 1\%
slower} than |stock|. The outliers (on either ends) were due to the difference
in the allocator and the GC between stock and Multicore OCaml.

The biggest impact was the increase in the code size due to stack overflow
checks. We define code size as the sum of the sizes of all the OCaml text
sections in the compiled binary file. The code size metric excludes data, debug
symbols, the runtime (implemented in C) and other C libraries. Compared to
|stock|, the code size for OCaml code is 19\% more for |MC| and |MC+RedZone32|,
and 30\% more for |MC+RedZone0|. The result shows that our 16-word red zone is
effective at reducing the code size compared to having no red zone, whereas the
32-word red zone does not noticeably reduce the code size further. Further work
is required to bring the code size closer to |stock|.

@@

\vspace{-3mm}
\subsection{No perform benchmarks}

@@

\vspace{-3mm}
\subsection{Concurrent benchmarks}

@@

\vspace{-2mm}
\section{Related Work}
\label{sec:related}

@@

The interaction of non-local control flow and resources has been studied
extensively. Scheme uses |dynamic-wind|~\cite{R5RS}, which is a generalisation of
Common Lisp |unwind-protect|~\cite{Steele90}, which ensures de-allocation (and
re-allocation) of resources every time the non-local control leaves and enters
back into a context. |dynamic-wind| is not quite sufficient for resource
clean up~\cite{Kiselyov,Sitaram03} as returning exits must be distinguished
from non-returning ones, and resource cleanup is only necessary during
non-returning exits.

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Sivaramakrishnan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.

\vspace{-2mm}
\section{Conclusions}
\label{sec:conc}

We have presented the design and implementation of effect handlers aimed at
retrofitting them onto the large body of existing OCaml code.  Our design walks
the tightrope of maintaining compatibility (e.g. for profiling and debuggers
and negligible overheads for existing programs) and also demonstrating good
performance and usability for new OCaml code that takes advantage of effect
handlers. Our evaluation shows that we have achieved our goal of balancing
compatibility with the introduction of effect handlers for mainstream OCaml
programming.

@@

Figures~\ref{res:macro_time} and \ref{res:macro_size} present the impact on
macro benchmarks that do not use effects using the Multicore OCaml compiler
which supports effect handlers. The aim is to quantify the impact of the
addition of effect handlers to the language on code that does not use effect
handlers. We define code size as the sum of the sizes of all the text sections
in the binary generated by compiling the OCaml code. Thus, the code size metric
excludes data, debug symbols, the runtime (implemented in C) and other
statically linked C libraries. The |MC| is Multicore OCaml with the default red
zone size (16 words). |MC+RedZone0| uses red zone size of 0; every OCaml
function has a stack overflow check. |MC+RedZone32| uses a red zone size of 32
words.
